### 2019-05-20 Writing/response_to_reviewers.tex

In the reviewer's demonstration, \texttt{x  = {1}} and \texttt{y  = {0}} thus become \texttt{x' = {1}} and \texttt{y' = {1}}.
Hence $CD_{(x',y')} = 0 \Leftrightarrow x' = y'$.
This is also reflected in the new $CD$ formula that makes comparing two characters for a single taxon nonsensical: the denominator is equal to $0$ if $n = 1$.

Furthermore, the CD metric, used in this manuscript as a proxy for character correlation actually measures the difference between in characters in terms of phylogenetic signal.
The character \texttt{X = {1, 0, 2, 1, 0}} described above implies two splits (and of course same goes for the translated \texttt{X' = {1, 2, 3, 1, 2}}).
Another character \texttt{Y = {1, 1, 1, 2, 2}} will imply only one split, not shared with \texttt{X} and a character \texttt{Z = {0, 2, 1, 0, 2}} (\texttt{Z' = {1, 2, 3, 1, 2}}) would imply the same split as \texttt{X} (hence, $CD_{(X',Z')} = 0$ and $CD_{(X',Y')} = 0.75$).
In other words the probability of X and Z generating conflicting split signal is null and this probability for X and Y is of 0.75.

Following these two points, the reviewer counter example $CD_{(x,y)} = 0$ means that 1) $x'$ and $y'$ are identical and the characters $x$ and $y$ imply the same number of splits in a tree (here no splits since only one taxon is used - however, this would hold for any $n$ number of taxa).
We have clarified these two points in the manuscript.

We've correct the supplementary materials 1 and added more explanations on (1) the character translation, (2) the four-point metric proof and (3) on how to interpret the $CD$ metric.
Additionally, we added more precisions in the text:

@@

\textit{(i.e. entirely correlated characters would define identical splits between a set of taxa)}. lines 199-200.

\item{\textcolor{blue}{While some details of the simulation study are included, some important ones are missing. In particular, in the design of the study, how did you avoid or mitigate errors introduced by the programs simulating data? All of the programs use random-number generators and have their own biases. How have you accounted for this in the study of design to conclude that your results are due to signal, and not bias from the programs used?}}

We do agree that generating true randomness in current software is tricky but we are not sure how to answer this comment.
The three first steps (generating the tree, the matrices and modifying the matrices) were using the default random number generator implemented in R, the Mersenne Twister algorithm that has a cycle of $2^{32}-1 = 4.2e^{9}$ unsigned random integers (see random numbers documentation in R: \texttt{?.Random.seed}).
The last step (generating the trees) was using both the random number generated by default my MrBayes and PAUP* based on the seeds from the different HPC we used (the Imperial College London Cluster Services and the Australian QRIScloud Awoonga cluster).

\begin{enumerate}
    \item \textbf{The generation of the ``true'' tree.}
    The generation of the true tree involves the drawing of two random parameters (birth and death parameters) and then builds the tree by drawing a probability of branching (birth) or extinction (death) at each time point until it reaches the desired number of taxa.
    This means at least $2 + 2tn$ where $t$ are the evaluated time points and $n$ the number of required taxa.
    For all 60 trees we thus have $120 + 500t$ draws. This means this whole step would cycle through all random numbers only if $t > 8.4e^{6}$. As a magnitude of comparisons when $t$ is one order of magnitude lower, $t = 1e^{5}$ it will typically lead to trees with $>> 1e{4}$ taxa (i.e. 2 orders of magnitude bigger than our biggest trees).

    \item \textbf{The generation of the discrete morphological characters (each 100, 350 and 1000).}
    The number of seeds involved in the generation of each model depends on the two models with a total of $5 + (2n-2) = 2n + 3$ random numbers for the HkY-binarised model (4 states proportions + 1 rate + number of branches) and $2 + (2n-2) = 2n$ for the Mk model (1 states numbers + 1 rate + number of branches).
    If if we simplify the seed calculations by saying only the HKY-model is chosen, we have at maximum $20(2n + 3)c$ random numbers; where $c$ is the number of characters.
    This results in a maximum of $1.4e{7}$ random numbers ($<< 4.2e^{9}$ total random numbers).

    \item \textbf{The removal of the characters (for the ``randomised'' matrices).}
    This step is also variable but we can simplify it by assuming the extreme case where 1) all characters are selected to be replaced and 2) they are all replaced by all characters. This results in a maximum of $20(5c)$ random numbers per matrix (2 for both replacements and 3 for each ``minimised'', ``maximised'' and ``randomised'' matrices).
    This results in a maximum of $1.4e{5}$ random numbers ($<< 4.2e^{9}$ total random numbers).
\end{enumerate}

Therefore, if cumulating all the seeds used in the generation of the matrices (trees + characters) and the modifications of the matrices is likely to be no more than $6.4e^{7}$.
Therefore we do not believe that the random number generation was a proper since we sampled two orders of magnitude less random numbers than the total ones available.

@@

\item{\textcolor{blue}{How did you ``randomly'' modify your matrices? Did you sample under some distribution? If so, which one?}}

Every replaced characters were randomly selected randomly from the total sample of characters (i.e. randomly sampling from the uniform integers distribution (1, $c$); where $c$ is the number of remaining characters). We have now specified this in the text:

\textit{Each replaced characters (i.e. the characters with a weight of +$1$) were randomly selected by sampling them randomly from the list of the remaining characters.} lines 296-297.

@@

\item{\textcolor{blue}{It is not obvious why generated character sequences are representative of biological sequences or sample the space overall. Were the parameters chosen for the HKY model to fit a particular evolutionary scenario? Are they meant to capture all of the space?}}

There is actually a vigorous debate surrounding how to generate biologically meaningful morphological characters \citep[][]{OReilly20160081,goloboff2017weighted,puttick2017uncertain,GoloboffEmpirical,OReillyEmpirical}.
This ongoing debate, however, focuses mainly on inference methods where data simulation is a tangential aspect of it.
Therefore, we decided to base our simulation protocol on the empirical observations showing that half of 206 analysed datasets supported a M$k$-like model \citep[][Fig. 3]{Wright01072016} and the other half being a combination of models not explicitly described here but that are believed to be covered with the HKY-binarised model \citep{puttick2017uncertain,OReillyEmpirical}.
In our simulations, the idea was to cover a broad range of morphological characters (with variable state numbers, evolution and transition rates) in order to capture a realistic image of the ``discrete morphological character space''.

@@

\item{\textcolor{blue}{There are statements that seem contradictory, like (line 335): ``Thus, additionally to the Wilcoxon test results, we considered distribution to be significantly similar if they had an overlap probability $>$ 0.95 and different if they had an overlap probability $>$ 0.05.''}}

We fixed this typo. We considered them different if they had an overlap probability $<$ 0.05.

\item{\textcolor{blue}{It is not clear how the conclusions were reached. For example, line 381-384: ``Regarding the inference method, there is a significant difference in clade conservation between Bayesian and maximum parsimony (Table 5 - median NTSRF of 0:828 and 0:679 respectively) but not in terms of individual taxon placements (Table 5 - median NTSTr of 0:738 and 0:601 respectively).'' Why is the 0:828 - 0:679 = 0:149 significant for this test and 0:738 - 0:601 = 0:137 is not?}}

The significances was referring to the results in Table 5 (Wilcoxon test $p$ value of 0 and 0.08 and Bhattacharrya Coefficient of 0.891 and 0.984 respectively) and the median values were extracted from the supplementary material 3 Table 5.
We now detailed the values and where they are present for each of the tests to avoid confusion.

### 2019-05-20 Writing/draft_highlighted.tex
reviewer 2 done with the random numbers detailed

\noindent Where $n$ is the number of taxa with comparable characters $x$, $y$ and $x_i$, $y_i$ are each character's state for the $i^{th}$ taxon.
%TG: Modifications in this sentence not in the letter yet
$CD$ is a continuous \hl{scaled Hamming distance} metric bounded between $0$ and $1$ (see the mathematical demonstration in the supplementary material 1).
Since we are considering differences as being only Fitch-like (non-additive) and unweighted, we calculated the difference between character states in a qualitative way.
Two same character states tokens have a difference of zero and two different ones have a difference of one (e.g. $0 - 0 = 0$ or $1 - 8 = 1$).
Additionally, we only consider differences for taxa with shared information \citep[i.e. a Gower distance;][]{GowerDist}.
\hl{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).}

We standardised each character by arbitrarily modifying their character state tokens (or symbols) by order of appearance.
In other words, we replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
\hl{This procedure allows to compare characters regardless of the significance given to their tokens (following the \textit{xyz} notation in }
\citealt{felsenstein2004inferring}; \hl{as used in} \citealt{Davalos01072014}).
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}}.
\hl{Similarly, an hypothetical comparison of two character for only one taxa \texttt{x = \{0\}} and \texttt{y = \{1\}} would result in comparing two identical characters \texttt{x' = \{1\}} and \texttt{y' = \{1\}}.}
Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct splits in the tree inference process).

When the character difference is null ($0$) it means that characters convey the same phylogenetic signal (i.e. characters are entirely correlated).
When the character difference is maximal ($1$) it means it conveys the greatest difference in phylogenetic signal (i.e. characters are uncorrelated).
It is important to stress that a character difference of $0$ (i.e. the same phylogenetic signal) does not mean the opposite of $1$ (i.e. \textit{not} the opposite phylogenetic signal but the most different number of implied splits) .
For example with three characters \texttt{A = \{0,1,1,1\}}, \texttt{B = \{1,0,0,0\}} and \texttt{C = \{0,1,2,3\}}, $CD_{(A,B)} = 0$ and $CD_{(A,C)} = 1$.
Because the character is continuous and bounded between $(0,1)$, it can be interpreted as the probability of two characters leading to a different set of splits (i.e. a different phylogenetic signal).

@@

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (i.e. with an average character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (i.e. with an average character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification (i.e. randomly weighting characters).
\hl{Each replaced characters (i.e. the characters with a weight of +$1$) where randomly selected from the remaining characters using a uniform distribution (0,1).}
Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of \hl{720} matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

>>>>

\noindent Where $n$ is the number of taxa with comparable characters $x$, $y$ and $x_i$, $y_i$ are each character's state for the $i^{th}$ taxon.
$CD$ is a continuous \hl{scaled Hamming distance} metric bounded between $0$ and $1$ (see the mathematical demonstration in the supplementary material 1).
Since we are considering differences as being only Fitch-like (non-additive) and unweighted, we calculated the difference between character states in a qualitative way.
Two same character states tokens have a difference of zero and two different ones have a difference of one (e.g. $0 - 0 = 0$ or $1 - 8 = 1$).
Additionally, we only consider differences for taxa with shared information \citep[i.e. a Gower distance;][]{GowerDist}.
\hl{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).}

We standardised each character by arbitrarily modifying their character state tokens (or symbols) by order of appearance.
In other words, we replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
\hl{This procedure allows to compare characters regardless of the significance given to their tokens (following the \textit{xyz} notation in }
\citealt{felsenstein2004inferring}; \hl{as used in} \citealt{Davalos01072014}).
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}}.
Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct splits in the tree inference process).
When the character difference is null ($0$) it means that characters convey the same phylogenetic signal (i.e. characters are entirely correlated).
When the character difference is maximal ($1$) it means it conveys the greatest difference in phylogenetic signal (i.e. characters are uncorrelated).
It is important to stress that a character difference of $0$ (i.e. the same phylogenetic signal) does not mean the opposite of $1$ (i.e. \textit{not} the opposite phylogenetic signal but the most different number of implied splits) .
For example with three characters \texttt{A = \{0,1,1,1\}}, \texttt{B = \{1,0,0,0\}} and \texttt{C = \{0,1,2,3\}}, $CD_{(A,B)} = 0$ and $CD_{(A,C)} = 1$.
Because the character is continuous and bounded between $(0,1)$, it can be interpreted as the probability of two characters leading to a different set of splits (i.e. a different phylogenetic signal).

@@

\subsection{Modifying the matrices}
We calculated the pairwise character differences for each generated matrix using the \texttt{dispRity R} package \citep{thomas_guillerme_2016_55646}.
We then modified the matrices to either maximise or minimise the pairwise character differences for each matrix using three different algorithms.
For maximising the pairwise differences between characters, we selected the characters that were the most similar to all the others (i.e. with an average character difference $<0.25$) and replaced them randomly by any of the remaining characters.
This operation increased the overall pairwise character difference in the matrix thus making the characters more dissimilar.
Conversely, for minimising the pairwise character differences, we selected the most dissimilar characters (i.e. with an average character difference $>0.75$) and randomly replaced them with the remaining ones.
Finally, because this operation effectively changes the weight of characters (i.e. giving the characters $<0.25$ or $>0.75$ a weight of $0$ and giving the randomly selected remaining characters a weight of +$1$), we randomly replaced the average number of characters replaced in the character maximisation and minimisation by any other characters as a randomised expectation modification (i.e. randomly weighting characters).
\hl{Each replaced characters (i.e. the characters with a weight of +$1$) were randomly selected by sampling them randomly from the list of the remaining characters.}
Each of the three matrices are effectively a bootstrap pseudo-replication of the ``normal'' matrix with the ``randomised'' one being a random one and the ``maximised'' and ``minimised'' being conditional bootstraps.
This step resulted in a total of \hl{720} matrices (hereafter called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices - see Fig. \ref{Fig:modif_matrix} for an illustration).
The algorithms for the three modifications are available on GitHub (\url{https://github.com/TGuillerme/CharactersCorrelation})

### 2019-05-20 Writing/supplementary/Sup1-CharDiff.tex
reviewer 2 done with the random numbers detailed

\subsection{Justification for the character states translation}
Because the character states tokens are not representing any meaningful numerical value if the characters are unordered (e.g. there is no assumption that state 0 should come before state 1), they need to be reordered in a standard way so that the first token in the character is renamed to ``1'', the second to ``2'' etc.
For example, in a 3 state character for four taxa: \texttt{X = {1, 0, 2, 1}}, the character tokens that are integers representing the states could equally have been \texttt{X = {E, A, I, E}} or \texttt{X = {blue, red, absent, blue}}.
Therefore, to compare characters fairly, the tokens needs to be translated in a standard way such as the first token is replaced by the symbol ``1'', etc.
In this example \texttt{X = {1, 0, 2, 1}} becomes \texttt{X' = {1, 2, 3, 1}}

\subsection{Interpretation}
For a characters \texttt{X = {1, 0, 2, 1}} implying three splits between four taxa and two characters \texttt{Y = {1, 1, 1, 2}} and \texttt{Z = {0, 2, 1, 0}} implying respectively two and three splits between four taxa, $CD_{(X,Y)} = 0.5$ and $CD_{(X,Z)} = 0$.
This can be interpreted as the probability of characters \texttt{X}, \texttt{Y} and \texttt{X}, \texttt{Z} giving a conflicting phylogenetic signal between the four taxa.
Here \texttt{X}, \texttt{Y} will conflict for 50\% of the taxa and \texttt{X}, \texttt{X} for 0\% of them.

@@

For any $n \in [1;\infty)$ and $x_{n}$ and $y_{n} \in (0,1)$ because the difference between character states is considered as Fitch-like (unordered; i.e. the difference between two different characters states is equal to $1$ and the difference between two identical character states is equal to $0$).
Furthermore, following the $xyz$ notation described above, for any $n > 1$, $x_{1}-y_{1}=0$.
Therefore the sum of the differences $\sum_{i=1}^{n}\abs{x_{i}-y_{i}} \in [0;(n-1)]$.

@@

For any $n \in [1;\infty)$ and $x_{n} = y_{n}$ then:
%TG: Is that valid for a demonstration?
\begin{equation}
    \label{identitydemo}
    \begin{aligned}
    \abs{x_{n}-y_{n}} &= 0\\
    \sum_{i}^{n}\abs{x_{i}-y_{i}} &= 0\\
    \frac{\sum_{i}^{n}\abs{x_{i}-y_{i}}}{n-1} &= 0\\
    \end{aligned}
\end{equation}

@@

For any $n \in [1;\infty)$ and $x_{n}$ and $y_{n} \in (0,1)$, similarly to equation \ref{positive_demo}:

@@

For any $n, \in [1;\infty)$ and $x_{n}$, $y_{n}$, $z_{n}$ $\in (0,1)$:

@@

For any $n \in [1;\infty)$ and any four distinct characters $i$, $j$, $n$ and $m$, and if equations \ref{Positivity} and \ref{Subadditivity} are true, the pairwise distances $CD_{(i,j)}$, $CD_{(m,n)}$, $CD_{(i,m)}$, $CD_{(j,n)}$, $CD_{(j,m)}$ and $CD_{(i,n)}$ can be represented as a combination of any five values $a$, $b$, $c$, $d$ and $e$ $\geq 0$:

>>>>

\subsection{Justification for the character states translation}
Because the character states tokens are not representing any meaningful numerical value if the characters are unordered (e.g. there is no assumption that state 0 should come before state 1), they need to be reordered in a standard way so that the first token in the character is renamed to ``1'', the second to ``2'' etc.
For example, in a 3 state character for four taxa: \texttt{X = {1, 0, 2, 1}}, the character tokens that are integers representing the states could equally have been \texttt{X = {E, A, I, E}} or \texttt{X = {blue, red, absent, blue}}.
Therefore, to compare characters fairly, the tokens needs to be translated in a standard way such as the first token is replaced by the symbol ``1'', etc.
In this example \texttt{X = {1, 0, 2, 1}} becomes \texttt{X' = {1, 2, 3, 1}}.

Note that the first two translated tokens are always equal (\texttt{x = {1, ...}} and \texttt{y = {1, ...}}) hence the denominator of $CD$ being $n - 1$ to reflect this equality.
Because of this, $CD$ can only be applied for $n \in [2;\infty)$ (i.e. if $n = 1$, the denominator is 0).

\subsection{Interpretation}
For a character \texttt{X = {1, 0, 2, 1, 0}} implying two splits between five taxa and two characters \texttt{Y = {1, 1, 1, 2, 2}} and \texttt{Z = {0, 2, 1, 0, 2}} implying respectively one and two splits between five taxa, $CD_{(X,Y)} = 0.75$ and $CD_{(X,Z)} = 0$.
This can be interpreted as the probability of characters \texttt{X}, \texttt{Y} and \texttt{X}, \texttt{Z} giving a conflicting phylogenetic signal between the four taxa.
Here \texttt{X}, \texttt{Y} will conflict for 70\% of the taxa and \texttt{X}, \texttt{X} for 0\% of them.

@@

For any $n \in [2;\infty)$ and $x_{n}$ and $y_{n} \in (0,1)$ because the difference between character states is considered as Fitch-like (unordered; i.e. the difference between two different characters states is equal to $1$ and the difference between two identical character states is equal to $0$).
Furthermore, following the $xyz$ notation described above, for any $n > 1$, $x_{1}-y_{1}=0$.
Therefore the sum of the differences $\sum_{i=1}^{n}\abs{x_{i}-y_{i}} \in [0;(n-1)]$.

@@

For any $n \in [2;\infty)$ and $x_{n} = y_{n}$ then:
%TG: Is that valid for a demonstration?
\begin{equation}
    \label{identitydemo}
    \begin{aligned}
    \abs{x_{n}-y_{n}} &= 0\\
    \sum_{i}^{n}\abs{x_{i}-y_{i}} &= 0\\
    \frac{\sum_{i}^{n}\abs{x_{i}-y_{i}}}{n-1} &= 0\\
    \end{aligned}
\end{equation}

@@

For any $n \in [2;\infty)$ and $x_{n}$ and $y_{n} \in (0,1)$, similarly to equation \ref{positive_demo}:

@@

For any $n, \in [2;\infty)$ and $x_{n}$, $y_{n}$, $z_{n}$ $\in (0,1)$:

@@

For any $n \in [2;\infty)$ and any four distinct characters $i$, $j$, $n$ and $m$, and if equations \ref{Positivity} and \ref{Subadditivity} are true, the pairwise distances $CD_{(i,j)}$, $CD_{(m,n)}$, $CD_{(i,m)}$, $CD_{(j,n)}$, $CD_{(j,m)}$ and $CD_{(i,n)}$ can be represented as a combination of any five values $a$, $b$, $c$, $d$ and $e$ $\geq 0$:
